Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-02-17 demystify.tex
updates

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
Originally, this book was written for international students, but has been expanded to include information for \textbf{US domestic students} (\autoref{chap:domestic-students}).
Moreover, while this primarily aims at students interested in CS, it might be relevant to students from various STEM (Science, Technologies, Engineering, and Mathematics) disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this book know about, the information should be generalizable to other good R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US.

@@

\chapter{Should You Apply?}\label{chap:should}

@@

\paragraph{Funding is not an issue} PhD students in CS \emph{do not} need to worry about funding, especially at good R1
universities in the US. 
If you are admitted, you will almost certainly
receive \emph{full funding} (\autoref{chap:funding}) to support your study. This includes tuition,
health insurance, and stipend (i.e., in STEM field you get paid for your study!). 
For many students, the stipends provided are not only sufficient to cover your own living expenses but also enough to support your families (e.g., spouse and children), allowing them to live comfortably.

@@

%While this handbook may not fully meet the needs of students aiming for top-10 schools (\autoref{sec:selecting-ranking-schools} and \autoref{chap:rankings}), which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other reputable R1 institutions.

%\begin{commentbox}[Vu:]
%    One of the reasons I created this document is that my colleagues at Mason are interested in recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (see \autoref{chap:ack}). Each year our CS PhD program receives 500+ PhD applications, most of which are international but only 5--6 are from Vietnam.
%  \end{commentbox}

@@

\emph{Funding}:  In many countries, funding comes from the university or the gov't. This funding often has a fixed duration, e.g., 3 or 4 years.  In the US (\autoref{chap:funding}), funding (e.g., RA) comes directly from your adviser (no fixed duration).  There are also fewer TA opportunities in European universities compared to the US.

@@

\chapter{How is Your Application Evaluated?}\label{chap:evalapps}

@@

\begin{commentbox}{Vu:}
At Mason, we usually decide that a full-time PhD candidate is either (i) admitted with funding (\autoref{chap:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding or reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision (\autoref{sec:ievaluate}) with a summary of your application, where we list
strengths, e.g., came from a well-known school, and weaknesses, e.g., weak and generic LoRs.
\end{commentbox}

@@

Some universities do waive, e.g., Rice, TTIC do not have fee for PhD applications and many universities for domestic students (\autoref{chap:domestic-students}).
Some programs also waive if the applicants attend some of their opening sessions.  Some programs waive if the applicant provide proof of financial difficulties, e.g., a statement from a financial adviser or a bank statement.

@@

% In CS, conferences are important and top conferences are prestigious. You can find top CS conferences at places such as \href{https://csrankings.org}{CSRankings}, which ranks CS programs based on how their faculty publish at top \emph{conferences} (see more in \autoref{chap:rankings}).

@@

While you might not have control over LORs (\autoref{chap:lor}) or where your go to school (\autoref{chap:your-school}), you do over your
statement of purpose (SOP) or personal statement\footnote{Few schools separate these documents and ask you to write both: SOP, which focuses on research experiences, and personal statement, which is everything more personal, e.g., why PhD, challenges, diversity, etc}! A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{chap:funding}).  Many SOP samples for CS are \href{https://cs-sop.org/}{available here}.

@@

\begin{itemize}
\item \textbf{Too personal} Don't talk about your personal issues, e.g., family, health, relationship, etc. This can raise concerns about your ability to handle the PhD. Don't talk about religious or political beliefs (\autoref{chap:cultural}).
\item \textbf{Criticizing} your current or previous institution or professors. Just like in a job interview, don't badmouth your current or previous employer. 
\item \textbf{No concrete evidence to back up claims}: e.g., saying you are passionate about a research topic without showing any experience. This is where specific names (work submitted at conf. X), numbers (outperformed SOTA by Y \%), and examples (worked on project Z) can help. 
\item \textbf{Use flowery and AI-like language}. Don't use AI to write your SOP. Not that hard to raise suspicion. Though you can ask AI to check your grammar and spelling.
\item \textbf{Not customized to the program}: If your SOP can be sent to multiple programs with few changes, it is too generic. Do some research and mention why you want to spend the next 5--7 years there.
\item \textbf{Mentioning wrong professors}: do not mention emeritus professors or those who have left.  Teaching and adjunct faculty are often not active in mentoring PhD students (\autoref{sec:faculty-types}). Do your homework and mention profs who are till active in research. 
\item \textbf{Too Long and fancy format}: As mentioned, keep it under 2 pages. Don't use too much coloring or fancy fonts (like those in Words). CS academics like using LaTeX (common way to write our papers and other documents), so write your SOP using LaTeX (with Times or default font, 11pt, and 1-inch margin).
\end{itemize}

@@

Unless your degrees are from the US or certain countries such as \href{https://www.gmu.edu/international/english-language-requirements}{these}, you will need to take standardized English tests. On one hand you will need to show some level of English proficiency, but on the other hand, you do not need to have very good scores in these tests (many adcom members themselves were once international students and struggled with English).
You should just do well enough to pass the minimum requirement set by the university. %, which is required for TA (\autoref{chap:funding}).
Just as with grades (\autoref{chap:your-school}) and GRE (\autoref{sec:gre}), having high scores in these tests might not help, but having too low scores can be a \red{red flag} and sometimes results in an automatic rejection (\autoref{sec:why-rejected}), e.g., below the minimum requirement.

@@

\section{CV}
\sectioninfo{Highlight and summarize major achievements such as major publications and competitive programming awards.}

@@

\part{After You Apply}\label{part:after-apply}
\chapter{Interview and the Waiting Game}\label{chap:interview}
\chapterinfo{Getting an interview is typically a good sign; but no interview does not mean rejection.}

@@

\chapter{Getting Admitted}\label{chap:accepted}
\chapterinfo{Congrats! Now it is your turn to evaluate the school!
Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now.}

@@

If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{chap:schoolsandprofs}), and funding availability (\autoref{chap:funding}). You will likely have to make your decision (\autoref{sec:accept-postpone-decline}) by around \emph{April 15}, which is the deadline for most schools
(\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why this date?}).

@@

\chapter{Dealing with Rejection}\label{chap:not-accepted}
\chapterinfo{Rejection is part of academia.  Common reasons for rejection include aiming too high, not a good fit, and bad luck.  Do not take rejection personally. Learn from it and try again.}

@@

\chapter{Funding}\label{chap:funding}
\chapterinfo{TAs, RAs, and fellowships are main funding sources for PhDs.  TAs are provided by the department to help with classes. RAs are given by profs. to help with their research.  Fellowships, provided by the university, department, or external sources such as government or industry, give move flexibility but can be very competitive.}

@@

  \tcblower
  Having health insurance is required at many US universities.  Do not assume that you're young and healthy and ignore insurance (\autoref{chap:cultural-misc}).  At Mason, and most CS PhD programs, your GTA or GRA comes with full insurance. In fact, at Mason your spouse/children will get a significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

For external fellowships including those from the US government (e.g., NSF GRFP, DOD NDSEG) and tech companies (e.g., Google, Microsoft, Facebook), you will need to apply.  % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.
Prestigious external fellowships typically require a clear and good research plan (the GRFP also requires broader impacts discussion). So it might be a good idea to wait until your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

@@

In addition to general fellowships that are available to all students, there are also specific fellowships, e.g., for US citizens and permanent residents (\autoref{chap:domestic-students}), for underrepresented minority (URM) students (\autoref{sec:urm}), and students from specific countries. There are many fellowships available, so you should look for them and apply.

In general, fellowships, especially those that are open to everyone, are highly competitive and prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only a few would have NSF Graduate (GRFP\footnote{\url{https://www.alexhunterlang.com/nsf-fellowship} is a good starting place for the GRFP with lots of proposal examples.}) or Microsoft fellowships. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even \emph{``honorable mentioning''}, your school will still mention you on their website, and you should put it on your CV. \autoref{chap:fellowships} discusses the evaluation processes of the NSF GRFP and DoD NDSEG.

@@

\chapter{Choosing Schools}\label{chap:schoolsandprofs}
\chapterinfo{Not every university has a CS PhD program. Not every professor, even those in CS, can advise or graduate CS PhD students.}

@@

Thus, if your goal is PhD in CS, you have to aim only for schools offering such a degree.  %This might sound obvious but many students (even those in the US) get confused due to the large number of universities.
While this can be confusing due to the large number of universities in the US, a little research, e.g., searching for PhD in CS from the school website, will help you find out. Schools listed in \autoref{chap:rankings} have PhD programs in CS, so you can start there.

@@

Instead of these superficial criteria, you should specifically consider the CS program and the research interests of faculty members.
You can learn about these using resources such as \href{https://csrankings.org}{CSRankings.org}. You will be very surprised to learn that a school that you didn't know much about can have very strong research in your topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty, e.g., who works on what, and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research findings (see \href{https://homes.cs.washington.edu/~mernst/advice/conferences-vs-journals.html}{why here}).}. %\autoref{chap:rankings} gives the top 50 CS programs in the US according to CSRankings.

@@

If you get admission to several places, you should consider attending Open Houses (\autoref{chap:accepted}) and contact profs. that you're interested in at those places and talk to them.  They would be more willing to chat with you now that you have been admitted.  Ask questions about \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Answers-to-Ph.D-Advisor-Guide}{their work}, how they manage students, and their expectations. You can even ask to contact their students.

@@

\emph{Teaching faculty}, sometimes called instructional faculty or professor of practice, mainly focuses on teaching. They typically teach 3--4 classes per semester (which is quite a lot) and \emph{do not} have research responsibilities, e.g., they do not have to worry much about publishing papers or obtaining grants. As such they typically do not have funding (e.g., RA~\autoref{chap:funding}) to support PhD students.
Depending on the university, they might be able to \emph{co-advise} and mentor PhD students with a tenure-line faculty, but they typically do not formally advise and graduate PhD students.

@@

First, if you want to contact a prof. to \emph{ask about your admission chance}, please \textbf{don't}. We don't know and can't answer because as explained in \autoref{chap:evalapps}, we don't make individual decisions and might not even be assigned to evaluate your application.  It is the same as sending a paper draft to a journal editor and asking them if your paper has a chance.

@@

\part{Miscs and FAQs}\label{part:faqs}
\chapter{Miscs and FAQs}\label{chap:faqs}

@@

Finally, I enter my decision, which is usually either a rejection or an offer of admission with full funding (e.g., from TA). I also recommend very strong candidates for the University Presidential fellowship., which is a fellowship from GMU that provides funding similar to an RA (\autoref{chap:funding}).
Note that while the system has other decision options, eg., admit without funding, provisional admission (e.g., if they need to take some courses), I do not use them, simply because we either reject or admit with funding.

@@

This is \emph{not necessary} and makes it look like you're trying to show off your connections. These nuances represent some cultural and academic differences that you may encounter and will gradually adapt to. More on these differences in \autoref{chap:cultural}.

@@

  \item[Adcom~\autoref{chap:evalapps}] Admission Committee. The group of faculty members who review applications and make admission decisions.

  \item [Adcom members~\autoref{chap:evalapps}] Faculty members who are part of the admission committee. These are the people who review applications and make admission decisions.

  \item[Adcom chair~\autoref{chap:evalapps}] The faculty member who leads the admission committee. This person is often not involved in individual admission decisions but oversees the process (e.g., resolve disputes, ensure fairness).

@@

  \item[International vs. Domestic students] International students are loosely defined as those who would need a visa to study in the US (\autoref{chap:visa}). Domestic students (\autoref{chap:domestic-students}) are those who do not need a visa, typically US citizens or permanent residents.

@@

  \item[Open House \autoref{chap:accepted}] An event where admitted students visit the university to meet faculty, students, and staff, and learn more about the program. This is a great opportunity to get a feel for the university and the department and decide if it is a good fit for you.

@@

\chapter{Domestic Students}\label{chap:domestic-students}
\chapterinfo{Specific benefits and opportunities for domestic students applying to CS PhD programs.}

@@

\paragraph{School \autoref{chap:your-school}} Adcom already knows about your school, which is a plus. You are also more familiar with the US education system and academic culture (\autoref{chap:cultural}).

@@

\paragraph{Funding \autoref{chap:funding}} You have more opportunities for funding, e.g., through government scholarships for US citizens and permanent residents.  You can also apply to specific programs \emph{before} you start your PhD, e.g., NSF Graduate Research Fellowship Program (GRFP) and Hertz Foundation Fellowship.  These fellowships are very competitive (\autoref{chap:fellowships}) and can significantly improve your admission chances.

@@

\paragraph{Open House \autoref{chap:accepted}} It is easier for you to attend open house events in person.  This can help you make a better decision on which school to attend.

@@

  \item Funding: MS is typically \emph{not funded}, while PhD is (\autoref{chap:funding}). See \autoref{sec:ms-funding} for more details on MS funding.

@@

\subsection{Funding}\label{sec:ms-funding}
Unlike PhD programs, which often have funding (\autoref{chap:funding}), MS students are typically \emph{not} funded. This is because MS students do not focus on research and thus are not funded through RA and PhD students have priority for TA positions.

@@

Because of the differences between MS and PhD programs (\autoref{sec:phd-vs-ms}), you should consider different factors when selecting and ranking MS programs. For example, you might want to consider factors such as location, industry connections, and job placement instead of research areas and advisers (\autoref{chap:schoolsandprofs}).

@@

\chapter{Fellowship Applications}\label{chap:fellowships}
\chapterinfo{How fellowships such as NSF GRFP and DOD DSEG are evaluated.}

Having an external and major fellowship can significantly improve your chances of getting into a good PhD program (\autoref{sec:fellowships}).  The National Science Foundation Graduate Research Fellowship Program (NSF GRFP) and the Department of Defense National Defense Science and Engineering Graduate Fellowship (DoD NDSEG) are two of the most prestigious fellowships for CS PhD students who are US citizens or permanent residents.  It is rumored that several universities would revert their admission decision on a student if they get one of these fellowships.

The \href{https://www.nsfgrfp.org/}{GRFP} and \href{https://www.ndsegfellowships.org/}{NDSEG} website have detailed information on the application process and eligibility requirements. This section focuses on how your application is evaluated and tips on improving your application. Note that while these are specific GRFP and NDSEG, the general advice can be applied to other major fellowships (those for non-US citizens or permanent residents).

%Securing a prestigious fellowship such as the National Science Foundation Graduate Research Fellowship Program (NSF GRFP) or the Department of Defense National Defense Science and Engineering Graduate Fellowship (DoD NDSEG) can significantly enhance a student’s prospects for admission to a strong PhD program. These fellowships not only provide financial support but also offer academic freedom and prestige that can set an applicant apart from their peers. However, successfully obtaining one of these fellowships is no easy feat—they are highly competitive, and applications are scrutinized rigorously by expert reviewers.

%One of the biggest advantages that applicants from elite institutions often have is access to faculty members who are well-versed in the review process. These faculty members understand the nuances of writing strong letters of recommendation (LoRs) and guiding students through the proposal-writing process. Unfortunately, students from smaller institutions or those with mentors unfamiliar with NSF or DoD fellowship reviews may find themselves at a disadvantage—not because of their potential or research ability, but simply due to a lack of insider knowledge about what makes a strong application.

%This chapter aims to level the playing field by providing an insider’s perspective on how these applications are reviewed. Drawing from my experience as a reviewer for both NSF GRFP and DoD NDSEG, I will demystify the evaluation process, highlight key elements that make applications stand out, and offer strategic advice for applicants and their letter writers. While research experience and proposal quality are, of course, crucial to success, understanding the review criteria and expectations can be just as important.

%Moreover, this chapter will provide guidance for LoR writers, helping them craft letters that effectively support their students’ applications. Many well-intentioned letters fail to address the aspects that reviewers prioritize, which can unintentionally weaken an otherwise strong application. By offering insights into what reviewers look for, I hope to empower both applicants and their mentors to present their strongest possible case for these fellowships.

%Whether you are an aspiring PhD student preparing your application or a faculty member writing a recommendation, this chapter will equip you with the knowledge needed to navigate the NSF GRFP and DoD NDSEG review process with confidence.

@@

The NSF GRFP is a prestigious fellowship for US citizens, nationals, and permanent residents who are in the early stages of their graduate study in STEM fields, including both MS and PhD students. The fellowship provides three years of financial support, which includes a stipend and tuition. The GRFP is highly competitive, with many applicants from top schools, with strong LoRs and research experience.

@@

\begin{itemize}
\item Each applicant is assigned to \emph{about 3 reviewers} who have expertise in the applicant's research topics. This means reviewers can be quite knowledgeable in your field and your work, your LoR writers and their work, and thus can evaluate your application quite in depth. This is different from PhD applications where adcom reviewers (\autoref{sec:adcom}) might not be in your field.

Note that if sometimes an original reviwer is no longer available then someone who is not in the field might review your application.  Thus you want to make your statements accessible to a general CS audience.

\item Reviewers typically try to compare applicants \emph{within their cohort level}.  This means that we do not compare an undergrad, who is still applying to grad school, to a 2nd year PhD student who already has an adviser and published multiple papers. This ensures that each applicant is evaluated fairly based on their level.

\item Unlike CS PhD applications that focus on research potentials in which papers, research experience, and LoRs are important, NSF GRFP looks at both \emph{research potential} and \emph{broader impacts}, e.g., how you can help society, how you can mentor others, etc.  It is not expected that you have published papers or have a lot of research experience (recall that MS students can also get the fellowship). Instead, you need to convince the reviewers that you have the potential to be a good researcher (technical merits) and that you can help society (broader impacts). More in \autoref{sec:grfp-im-bi}.

@@

\subsection{Intellectual Merit and Broader Impacts}\label{sec:grfp-im-bi}

Your application will be evaluated based on two criteria: Intellectual Merit (IM) and Broader Impacts (BI).

>>>>

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
Originally, this book was intended for international students, but has been expanded to include information for \textbf{US domestic students} (\autoref{sec:domestic-students}).
Moreover, while this primarily aims at students interested in CS, it might be relevant to students from various STEM (Science, Technologies, Engineering, and Mathematics) disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this book know about, the information should be generalizable to other good R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US.

@@

\chapter{Should You Apply?}\label{sec:should}

@@

\paragraph{Funding is not an issue} PhD students in CS \emph{do not} need to worry about funding, especially at good R1
universities in the US. 
If you are admitted, you will almost certainly
receive \emph{full funding} (\autoref{sec:funding}) to support your study. This includes tuition,
health insurance, and stipend (i.e., in STEM field you get paid for your study!). 
For many students, the stipends provided are not only sufficient to cover your own living expenses but also enough to support your families (e.g., spouse and children), allowing them to live comfortably.

@@

%While this handbook may not fully meet the needs of students aiming for top-10 schools (\autoref{sec:selecting-ranking-schools} and \autoref{sec:ranking}), which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other reputable R1 institutions.

%\begin{commentbox}[Vu:]
%    One of the reasons I created this document is that my colleagues at Mason are interested in recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (see \autoref{sec:ack}). Each year our CS PhD program receives 500+ PhD applications, most of which are international but only 5--6 are from Vietnam.
%  \end{commentbox}

@@

\emph{Funding}:  In many countries, funding comes from the university or the gov't. This funding often has a fixed duration, e.g., 3 or 4 years.  In the US (\autoref{sec:funding}), funding (e.g., RA) comes directly from your adviser (no fixed duration).  There are also fewer TA opportunities in European universities compared to the US.

@@

\chapter{How is Your Application Evaluated?}\label{sec:evalapps}

@@

\begin{commentbox}{Vu:}
At Mason, we usually decide that a full-time PhD candidate is either (i) admitted with funding (\autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding or reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision (\autoref{sec:ievaluate}) with a summary of your application, where we list
strengths, e.g., came from a well-known school, and weaknesses, e.g., weak and generic LoRs.
\end{commentbox}

@@

Some universities do waive, e.g., Rice, TTIC do not have fee for PhD applications and many universities for domestic students (\autoref{sec:domestic-students}).
Some programs also waive if the applicants attend some of their opening sessions.  Some programs waive if the applicant provide proof of financial difficulties, e.g., a statement from a financial adviser or a bank statement.

@@

% In CS, conferences are important and top conferences are prestigious. You can find top CS conferences at places such as \href{https://csrankings.org}{CSRankings}, which ranks CS programs based on how their faculty publish at top \emph{conferences} (see more in \autoref{sec:ranking}).

@@

While you might not have control over LORs (\autoref{chap:lor}) or where your go to school (\autoref{chap:your-school}), you do over your
statement of purpose (SOP) or personal statement\footnote{Few schools separate these documents and ask you to write both: SOP, which focuses on research experiences, and personal statement, which is everything more personal, e.g., why PhD, challenges, diversity, etc}! A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{sec:funding}).  Many SOP samples for CS are \href{https://cs-sop.org/}{available here}.

@@

\begin{itemize}
\item \textbf{Too personal} Don't talk about your personal issues, e.g., family, health, relationship, etc. This can raise concerns about your ability to handle the PhD. Don't talk about religious or political beliefs (\autoref{sec:cultural}).
\item \textbf{Criticizing} your current or previous institution or professors. Just like in a job interview, don't badmouth your current or previous employer. 
\item \textbf{No concrete evidence to back up claims}: e.g., saying you are passionate about a research topic without showing any experience. This is where specific names (work submitted at conf. X), numbers (outperformed SOTA by Y \%), and examples (worked on project Z) can help. 
\item \textbf{Use flowery and AI-like language}. Don't use AI to write your SOP. Not that hard to raise suspicion. Though you can ask AI to check your grammar and spelling.
\item \textbf{Not customized to the program}: If your SOP can be sent to multiple programs with few changes, it is too generic. Do some research and mention why you want to spend the next 5--7 years there.
\item \textbf{Mentioning wrong professors}: do not mention emeritus professors or those who have left.  Teaching and adjunct faculty are often not active in mentoring PhD students (\autoref{sec:faculty-types}). Do your homework and mention profs who are till active in research. 
\item \textbf{Too Long and fancy format}: As mentioned, keep it under 2 pages. Don't use too much coloring or fancy fonts (like those in Words). CS academics like using LaTeX (common way to write our papers and other documents), so write your SOP using LaTeX (with Times or default font, 11pt, and 1-inch margin).
\end{itemize}

@@

Unless your degrees are from the US or certain countries such as \href{https://www.gmu.edu/international/english-language-requirements}{these}, you will need to take standardized English tests. On one hand you will need to show some level of English proficiency, but on the other hand, you do not need to have very good scores in these tests (many adcom members themselves were once international students and struggled with English).
You should just do well enough to pass the minimum requirement set by the university. %, which is required for TA (\autoref{sec:funding}).
Just as with grades (\autoref{chap:your-school}) and GRE (\autoref{sec:gre}), having high scores in these tests might not help, but having too low scores can be a \red{red flag} and sometimes results in an automatic rejection (\autoref{sec:why-rejected}), e.g., below the minimum requirement.

@@

\section{CV}
\sectioninfo{Highlight and summarize major achievements, e.g., Publications, Programming Awards}

@@

\part{After You Apply}\label{part:after-apply}
\chapter{Interview and the Waiting Game}\label{chap:interview}
\chapterinfo{Getting an interview is typically a good sign; but no interview does not mean rejection}

@@

\chapter{Getting Admitted}\label{sec:accepted}
\chapterinfo{Congrats! Now it is your turn to evaluate the school!
Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now.}

@@

If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will likely have to make your decision (\autoref{sec:accept-postpone-decline}) by around \emph{April 15}, which is the deadline for most schools
(\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why this date?}).

@@

\chapter{Dealing with Rejection}\label{chap:not-accepted}
\chapterinfo{Rejection is common and part of academia.  Common reasons for rejection include aiming too high, not a good fit, or bad luck. Learn from it and try again.}

@@

\chapter{Funding}\label{sec:funding}
\chapterinfo{TAs, RAs, and fellowships are main funding sources for PhDs.  TAs are provided by the department to help with classes. RAs are given by profs. to help with their research.  Fellowships, provided by the university, department, or external sources such as government or industry, give move flexibility but can be very competitive.}

@@

  \tcblower
  Having health insurance is required at many US universities.  Do not assume that you're young and healthy and ignore insurance (\autoref{sec:cultural-misc}).  At Mason, and most CS PhD programs, your GTA or GRA comes with full insurance. In fact, at Mason your spouse/children will get a significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

For external fellowships including those from the US government (e.g., NSF, DOD) and tech companies (e.g., Google, Microsoft, Facebook), you will need to apply.  % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.
Prestigious external fellowships typically require a clear and good research plan (the GRFP also requires outreach activities and plans). So it might be a good idea to wait until your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

@@

In addition to general fellowships that are available to all students, there are also specific fellowships, e.g., for US citizens and permanent residents (\autoref{sec:domestic-students}), for underrepresented minority (URM) students (\autoref{sec:urm}), and students from specific countries. There are many fellowships available, so you should look for them and apply.

In general, fellowships, especially those that are open to everyone, are highly competitive and prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only a few would have NSF Graduate (GRFP\footnote{\url{https://www.alexhunterlang.com/nsf-fellowship} is a good starting place for the GRFP with lots of proposal examples.}) or Microsoft fellowships. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even \emph{``honorable mentioning''}, your school will still mention you on their website, and you should put it on your CV.

@@

\chapter{Choosing Schools}\label{sec:schoolsandprofs}
\chapterinfo{Not every university has a CS PhD program. Not every professor, even those in CS, can advise or graduate CS PhD students.}

@@

Thus, if your goal is PhD in CS, you have to aim only for schools offering such a degree.  %This might sound obvious but many students (even those in the US) get confused due to the large number of universities.
While this can be confusing due to the large number of universities in the US, a little research, e.g., searching for PhD in CS from the school website, will help you find out. Schools listed in \autoref{sec:ranking} have PhD programs in CS, so you can start there.

@@

Instead of these superficial criteria, you should specifically consider the CS program and the research interests of faculty members.
You can learn about these using resources such as \href{https://csrankings.org}{CSRankings.org}. You will be very surprised to learn that a school that you didn't know much about can have very strong research in your topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty, e.g., who works on what, and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research findings (see \href{https://homes.cs.washington.edu/~mernst/advice/conferences-vs-journals.html}{why here}).}. %\autoref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

If you get admission to several places, you should consider attending Open Houses (\autoref{sec:accepted}) and contact profs. that you're interested in at those places and talk to them.  They would be more willing to chat with you now that you have been admitted.  Ask questions about \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Answers-to-Ph.D-Advisor-Guide}{their work}, how they manage students, and their expectations. You can even ask to contact their students.

@@

\emph{Teaching faculty}, sometimes called instructional faculty or professor of practice, mainly focuses on teaching. They typically teach 3--4 classes per semester (which is quite a lot) and \emph{do not} have research responsibilities, e.g., they do not have to worry much about publishing papers or obtaining grants. As such they typically do not have funding (e.g., RA~\autoref{sec:funding}) to support PhD students.
Depending on the university, they might be able to \emph{co-advise} and mentor PhD students with a tenure-line faculty, but they typically do not formally advise and graduate PhD students.

@@

First, if you want to contact a prof. to \emph{ask about your admission chance}, please \textbf{don't}. We don't know and can't answer because as explained in \autoref{sec:evalapps}, we don't make individual decisions and might not even be assigned to evaluate your application.  It is the same as sending a paper draft to a journal editor and asking them if your paper has a chance.

@@

\part{Miscs and FAQs}\label{part:faqs}
\chapter{Miscs and FAQs}\label{sec:faqs}

@@

Finally, I enter my decision, which is usually either a rejection or an offer of admission with full funding (e.g., from TA). I also recommend very strong candidates for the University Presidential fellowship., which is a fellowship from GMU that provides funding similar to an RA (\autoref{sec:funding}).
Note that while the system has other decision options, eg., admit without funding, provisional admission (e.g., if they need to take some courses), I do not use them, simply because we either reject or admit with funding.

@@

This is \emph{not necessary} and makes it look like you're trying to show off your connections. These nuances represent some cultural and academic differences that you may encounter and will gradually adapt to. More on these differences in \autoref{sec:cultural}.

@@

  \item[Adcom~\autoref{sec:evalapps}] Admission Committee. The group of faculty members who review applications and make admission decisions.

  \item [Adcom members~\autoref{sec:evalapps}] Faculty members who are part of the admission committee. These are the people who review applications and make admission decisions.

  \item[Adcom chair~\autoref{sec:evalapps}] The faculty member who leads the admission committee. This person is often not involved in individual admission decisions but oversees the process (e.g., resolve disputes, ensure fairness).

@@

  \item[International vs. Domestic students] International students are loosely defined as those who would need a visa to study in the US (\autoref{chap:visa}). Domestic students (\autoref{sec:domestic-students}) are those who do not need a visa, typically US citizens or permanent residents.

@@

  \item[Open House \autoref{sec:accepted}] An event where admitted students visit the university to meet faculty, students, and staff, and learn more about the program. This is a great opportunity to get a feel for the university and the department and decide if it is a good fit for you.

@@

\chapter{Domestic Students}\label{sec:domestic-students}
\chapterinfo{Specific benefits and opportunities for domestic students applying to CS PhD programs.}

@@

\paragraph{School \autoref{chap:your-school}} Adcom already knows about your school, which is a plus. You are also more familiar with the US education system and academic culture (\autoref{sec:cultural}).

@@

\paragraph{Funding \autoref{sec:funding}} You have more opportunities for funding, e.g., through government scholarships for US citizens and permanent residents.  You can also apply to specific programs \emph{before} you start your PhD, e.g., NSF Graduate Research Fellowship Program (GRFP) and Hertz Foundation Fellowship.  These fellowships are very competitive and can significantly improve your admission chances.

@@

\paragraph{Open House \autoref{sec:accepted}} It is easier for you to attend open house events in person.  This can help you make a better decision on which school to attend.

@@

  \item Funding: MS is typically \emph{not funded}, while PhD is (\autoref{sec:funding}). See \autoref{sec:ms-funding} for more details on MS funding.

@@

\subsection{Funding}\label{sec:ms-funding}
Unlike PhD programs, which often have funding (\autoref{sec:funding}), MS students are typically \emph{not} funded. This is because MS students do not focus on research and thus are not funded through RA and PhD students have priority for TA positions.

@@

Because of the differences between MS and PhD programs (\autoref{sec:phd-vs-ms}), you should consider different factors when selecting and ranking MS programs. For example, you might want to consider factors such as location, industry connections, and job placement instead of research areas and advisers (\autoref{sec:schoolsandprofs}).

@@

\chapter{Fellowship Applications}\label{sec:fellowship}
\chapterinfo{How to apply for fellowships such as NSF GRFP, DOD DSEG to fund your PhD.}

@@

@@

\begin{itemize}
\item Each applicant is assigned to \emph{about 3 reviewers} who all seem to have expertise in the applicant's research topics. This means reviewers can be quite knowledgeable in your field and your work, your LoR writers and their work, and thus can evaluate your application quite in depth. This is different from PhD applications where adcom reviewers (\autoref{sec:adcom}) might not be in your field.

Note that if sometimes an original reviwer is no longer available then someone who is not in the field might review your application.  Thus in general you want to make your application accessible to a general audience in CS.

\item Reviewers attempt to compare applicants \emph{within their cohort level}.  This means that we do not compare an undergrad, who is still applying to grad school, to a 2nd year PhD student who already has an adviser and published multiple papers. This ensures that each applicant is evaluated fairly based on their level.

\item Unlike CS PhD applications, which focus on research potentials in which concrete evidence like papers, research experience, and LoRs are important, NSF GRFP also focuses on \emph{potential} but also on broader impacts, e.g., how you can help society, how you can mentor others, etc.

@@

\subsection{Intellectual Merit and Broader Impacts}
Your will be evaluated based on two criteria: Intellectual Merit (IM) and Broader Impacts (BI).

\paragraph{Intellectual merit} is based on your research potential, e.g., your research experience, your potential to do research, and how you can contribute to the field.
